---
title: 'R project: Data analysis and visualization of a publicly available dataset'
author: "Julie Pevernagie"
date: "January 2019"
output:
  pdf_document: default
  word_document: default
  html_document: default
header-includes:
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, error=FALSE, message=FALSE)
```

# Human Freedom Index
## About the dataset

The Human Freedom Index presents the state of human freedom in the world based 
on a broad measure that encompasses personal, civil, and economic freedom. Human 
freedom is a social concept that recognizes the dignity of individuals and is 
defined here as negative liberty or the absence of coercive constraint. Because 
freedom is inherently valuable and plays a role in human progress, it is worth 
measuring carefully. The Human Freedom Index is a resource that can help to more 
objectively observe relationships between freedom and other social and economic 
phenomena, as well as the ways in which the various dimensions of freedom 
interact with one another.

This dataset uses 79 distinct indicators of personal and economic freedom in the 
following areas:


**Personal Freedom**

* Rule of Law
* Security and Safety
* Movement
* Religion
* Association, Assembly, and Civil Society
* Expression and Information
* Identity and Relationships


**Economical Freedom**

* Size of Government
* Legal System and Property Rights
* Access to Sound Money
* Freedom to Trade Internationally
* Regulation of Credit, Labor, and Business


The HFI is the most comprehensive freedom index so far created for a globally 
meaningful set of countries. The HFI covers 162 countries for 2016, the most 
recent year for which sufficient data are available and it includes three 
countries - Belarus, Iraq, and Sudan - that were added this year. The index 
ranks countries beginning in 2008, the earliest year for which a robust enough 
index could be produced.

On a scale of 0 to 10, where 10 represents more freedom, the average human 
freedom rating for 162 countries in 2016 was 6.89. Among countries included in 
this year’s and last year’s report, the level of freedom decreased slightly 
(-0.01) compared with 2015, with 63 countries increasing their ratings and 87 
decreasing. Since 2008, the level of global freedom has also decreased slightly 
(-0.06), with 56 countries in the index increasing their ratings and 81 
decreasing.

\newpage
### Loading the dataset and required packages:

```{r echo=TRUE, results='hide'}
# check installed pkg
pkg <- installed.packages()[, "Package"]
loadpkg <- c("ggplot2","rworldmap", "reshape2","RColorBrewer","ggplus") 

library(devtools)
devtools::install_github("guiastrennec/ggplus")

# install pkg if necessary
for ( i in 1:length(loadpkg) ){
  if(!(loadpkg[i] %in% pkg)) {
    install.packages(loadpkg[i]) 
  }
}

# load the required packages
lapply(loadpkg, library, character.only = TRUE)

# load data for global human freedom index 2008-2016
url <- "https://github.com/Juliepvr/hello-world/raw/master/hfi_cc_2018.csv"
hfi <- read.csv(url)

```

### Modifying the data for easy processing:
```{r}

hfi$year<- as.factor(hfi$year)

# data for Belgium:
hfiBEL <- hfi[hfi$ISO_code=="BEL",]

# data for 2016:
### global:
hfi2016 <- hfi[hfi$year==2016,]
### for Belgium:
hfiBEL2016<- hfiBEL[hfiBEL$year==2016,]

# load seperate text files with titles
titles <- scan(
  "https://raw.githubusercontent.com/Juliepvr/hello-world/master/hfi-titles", 
  what="", sep="\n")
subt <- scan(
  "https://raw.githubusercontent.com/Juliepvr/hello-world/master/hfi-list-colum-names", 
  what="", sep="\n")

# execute R commands in textfile line by line:
subtitles<- c()
for(i in 1:length(subt)){
  tmp<-list(eval(parse(text=subt[i])))
  subtitles<-append(subtitles, tmp, after=length(subtitles))
}

# Filter out columns to be used from BEL 2016
skip_this <- grep( ("((\\_([a-zA-Z]{1,})){3,})"), colnames(hfiBEL2016), 
                   value=TRUE) 
hfiBEL2016 <- hfiBEL2016[,!(colnames(hfiBEL2016) %in% skip_this)]

# totals and subcategories seperate
seperate_this<-grep( ("((\\_([a-zA-Z]{1,})){2})"), colnames(hfiBEL2016), 
                     value=TRUE)
totalsBEL<-hfiBEL2016[,!(colnames(hfiBEL2016) %in% seperate_this)]
totalsBEL<-totalsBEL[,c(5:11,14:18)] # cut out pf and ef totals and rank
cols<-colnames(totalsBEL)
detailsBEL<-hfiBEL2016[,seperate_this]

for(i in 1:length(cols)){
  assign(cols[i], detailsBEL[,grep(paste(cols[i], "_*", sep = ""), 
                                   colnames(detailsBEL))])
}

# picking a color palette for uniform plotting, custom colors for better 
# representation of the data.

clr <- c("black",brewer.pal(11,"RdGy")[c(11,10,1)], brewer.pal(6,"RdYlGn"))

```
\newpage
\blandscape

## Worldwide Freedom 2016

``` {r fig.height=7, fig.width=10.5}

# Worldmap with totals for 2016

# create SpatialPolygonsDataFrame: match countries between data and package
sPDF <- joinCountryData2Map( hfi2016
                             , joinCode = "ISO3"
                             , nameJoinColumn = "ISO_code")
                             
# draw map
mapParams <- mapCountryData(sPDF, nameColumnToPlot="hf_score", 
                            mapTitle="Human Freedom Index 2016",
                            catMethod=0:10,
                            colourPalette=clr, 
                            addLegend=FALSE )
# draw legend
do.call( addMapLegend, 
         c(mapParams, 
           legendLabels="all", 
           legendWidth=0.5, 
           legendIntervals="page", 
           legendMar = 4 ))

```
\newpage

## Distribution of Freedom by Region

The violin plots show that western countries enjoy more freedom and the scores 
are more uniform amongst the countries, as opposed to the Middle East with 
pronounced outliers. Since North America consist of only 2 countries (Canada and 
USA), there is not enough data to generate a violinplot.
```{r fig.height=7, fig.width=10.5}
# Violin plots + boxplots per region

viobox <- ggplot(hfi2016, aes(x=hfi2016$region, y=hfi2016$hf_score, 
                              fill=hfi2016$region)) +   
  geom_violin(trim=FALSE,show.legend = FALSE) + 
  geom_boxplot(width=0.1, fill="white") +
  labs(title="Human Freedom by Region",x="Region", y="Human Freedom") 

viobox + theme_light() + theme(axis.text.x=element_text(angle=45, hjust=1), 
                            plot.margin=margin(1, 1, 1, 2, "cm"),
                            plot.title = element_text(hjust = 0.5)) 
```
\elandscape
\newpage

## Belgium

The following graph illustrates the evolution of personal freedom, economical 
freedom and the resulting human freedom in Belgium since 2008. 
```{r fig.height=5, fig.width=7.5}
#  Overview Belgium 2008-2016

summBEL <-data.frame(hfiBEL$year,hfiBEL$pf_score,hfiBEL$ef_score,hfiBEL$hf_score)
evolBEL <- data.frame(hfiBEL$year,hfiBEL$hf_score)
summBEL <- melt(summBEL)
evolBEL <- melt(evolBEL)
levels(summBEL$variable)<- c("Personal Freedom","Economical Freedom",
                             "Human Freedom")

a<-round(mean(hfiBEL$pf_score, digit=0))
b<-round(mean(hfiBEL$ef_score, digit=0))
c<-round(mean(hfiBEL$hf_score, digit=0))

p1 <- ggplot(summBEL, aes(summBEL$hfiBEL.year,summBEL$value,
                         fill = summBEL$variable)) 

# graph evolution areas of freedom
p1 + geom_line(aes(color = summBEL$variable),group = summBEL$variable, size=2)+ 
  labs(title="Evolution of Freedom in Belgium",x="Year", y="Score") +
  guides(color=guide_legend("Legend")) + 
  scale_color_manual(values=clr[c(a,b,c)]) + 
  theme_light() + 
  theme(plot.margin=margin(1, 1, 1, 1, "cm"), 
        plot.title = element_text(hjust = 0.5))+
  annotate("text", x=.75, y=c(a,b,c)+.2,label = c("PF","EF","HF"), 
           color=clr[c(a,b,c)])

```

\newpage

## Details

```{r fig.height=10.5, fig.width=7}
# Details Belgium
BELdata <- list()

for(i in 1:length(cols)) {
  BELdata[[i]] <- eval(parse(text=cols[i]))
}

BELdata<-melt(BELdata)
BELdata$variable<- factor(BELdata$variable, levels = BELdata$variable, labels = 
                           unlist(subtitles)) 
BELdata$L1<- factor(BELdata$L1, levels = 1:12, labels = titles)

p2 <-ggplot(BELdata, value.name=titles,aes(x=variable,y=value, fill=value))+
    geom_col(position="dodge")+
    ylim(0,10) +
    coord_flip()+
    scale_fill_gradient2( midpoint=5, low=clr[5], mid=clr[7],high=clr[9]) +
    labs(title="Areas of Freedom for Belgium in 2016",x=element_blank(), 
          y=element_blank()) +
    theme_light() +
    theme(plot.title = element_text(hjust = 0.5),legend.position="none")


# Plot on multiple pages (ggplus package from github)
facet_multiple(plot = p2, 
              facets = "L1", 
              scales = "free",
              ncol = 1, 
              nrow = 4)
```

\newpage
\blandscape

## Women's freedom of movement

One of the components of the human freedom index measures the freedom of women 
to move outside the home. The following elements were considered: ability to 
choose their places of residence, to visit their families and friends, or to 
apply for a passport. A rating of 10 was assigned to countries with no 
restrictions on women’s movement outside the home. A rating of 5 was assigned to 
countries where (some) women can leave home sometimes but with restrictions. A 
rating of 0 was assigned to countries where women can never leave home without 
restrictions (e.g., they need a male companion).

``` {r fig.height=7, fig.width=10.5}

# map women

hfi2016$pf_movement_women <- as.factor(hfi2016$pf_movement_women)
levels(hfi2016$pf_movement_women)<-c("none","limited","all")

# create SpatialPolygonsDataFrame: match countries between data and package
sPDF <- joinCountryData2Map( hfi2016
                             , joinCode = "ISO3"
                             , nameJoinColumn = "ISO_code")
                             
# draw map
mapParams <- mapCountryData(sPDF, nameColumnToPlot="pf_movement_women", 
                            mapTitle="Women's freedom by law to move outside the 
                            home 2016", 
                            catMethod="categorical",
                            colourPalette=clr[c(5,7,9)])

```
\elandscape