---
title: "R project: Data analysis and visualization of a publicly available dataset"
author: "Julie Pevernagie"
date: "January 2019"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, error=FALSE, message=FALSE)
```


# Human Freedom Index
## About the dataset

The Human Freedom Index presents the state of human freedom in the world based 
on a broad measure that encompasses personal, civil, and economic freedom. Human 
freedom is a social concept that recognizes the dignity of individuals and is 
defined here as negative liberty or the absence of coercive constraint. Because 
freedom is inherently valuable and plays a role in human progress, it is worth 
measuring carefully. The Human Freedom Index is a resource that can help to more 
objectively observe relationships between freedom and other social and economic 
phenomena, as well as the ways in which the various dimensions of freedom 
interact with one another.

This dataset uses 79 distinct indicators of personal and economic freedom in the 
following areas:


**Personal Freedom**

* Rule of Law
* Security and Safety
* Movement
* Religion
* Association, Assembly, and Civil Society
* Expression and Information
* Identity and Relationships


**Economical Freedom**

* Size of Government
* Legal System and Property Rights
* Access to Sound Money
* Freedom to Trade Internationally
* Regulation of Credit, Labor, and Business


The HFI is the most comprehensive freedom index so far created for a globally 
meaningful set of countries. The HFI covers 162 countries for 2016, the most 
recent year for which sufficient data are available and it includes three 
countries - Belarus, Iraq, and Sudan - that were added this year. The index 
ranks countries beginning in 2008, the earliest year for which a robust enough 
index could be produced.

On a scale of 0 to 10, where 10 represents more freedom, the average human 
freedom rating for 162 countries in 2016 was 6.89. Among countries included in 
this year’s and last year’s report, the level of freedom decreased slightly 
(-0.01) compared with 2015, with 63 countries increasing their ratings and 87 
decreasing. Since 2008, the level of global freedom has also decreased slightly 
(-0.06), with 56 countries in the index increasing their ratings and 81 
decreasing.

### Loading the dataset and required packages:

``` {r} 
# check installed pkg
pkg <- installed.packages()[, "Package"]
loadpkg <- c("ggplot2","rworldmap", "reshape2","RColorBrewer","ggplus") 

#install.packages("devtools")
#library(devtools)
#devtools::install_github("guiastrennec/ggplus")

# install pkg if necessary
for ( i in 1:length(loadpkg) ){
  if(!(loadpkg[i] %in% pkg)) { 
    install.packages(loadpkg[i]) 
  }
}

# load the required packages
lapply(loadpkg, library, character.only = TRUE)

# load data for global human freedom index 2008-2016
url <- "https://github.com/Juliepvr/R-project/raw/master/hfi_cc_2018.csv"
hfi <- read.csv(url)

# load seperate text files with titles
titles <- scan(
  "https://raw.githubusercontent.com/Juliepvr/R-project/master/hfi-titles", 
  what="", sep="\n")
subt <- scan(
  "https://raw.githubusercontent.com/Juliepvr/R-project/master/hfi-list-colum-names", 
  what="", sep="\n")
# execute R commands in the textfile line by line:
subtitles<- c()
for(i in 1:length(subt)){
  tmp<-list(eval(parse(text=subt[i])))
  subtitles<-append(subtitles, tmp, after=length(subtitles))
}

clr<-c("black",brewer.pal(11,"RdGy")[c(11,10,1)] , brewer.pal(6,"RdYlGn"))

hfi$year<- as.factor(hfi$year)
hfi$countries <- as.factor(hfi$countries)

hfi_summary <- hfi[,c("year", "ISO_code", "countries","region","pf_score", "ef_score", "hf_score")]
colnames(hfi_summary)<-c("year", "ISO_code", "countries","region","Personal Freedom","Economical Freedom","Human Freedom")

# create subset
skip_this <- grep( ("((\\_([a-zA-Z]{1,})){3,})"), colnames(hfi), 
                   value=TRUE) 
hfi <- hfi[,!(colnames(hfi) %in% skip_this)]

# totals and subcategories seperate
seperate_this<-grep( ("((\\_([a-zA-Z]{1,})){2})"), colnames(hfi), 
                     value=TRUE)
totals<-hfi[,!(colnames(hfi) %in% seperate_this)]
totals<-totals[,c(5:11,14:18)] # cut out pf and ef totals and rank
cols<-colnames(totals)
colnames(totals)<-titles


```

## Worldmap 

```{r}

# Choose year, choose column
inputPanel(
  selectInput("yr", label = "Select Year",
              choices = levels(hfi$year), selected = levels(hfi$year)[length(levels(hfi$year))]),
  
  selectInput("frdm", label = "Select Area of freedom",
              choices = colnames(hfi_summary)[5:7] , selected = "Human Freedom")
)

renderPlot({
  hfiYR <- hfi[hfi$year==input$yr,]
  hfi_summYR <- hfi_summary[hfi_summary$year==input$yr,]
  
  
  # create SpatialPolygonsDataFrame: match countries between data and package
sPDF <- joinCountryData2Map( hfi_summYR
                             , joinCode = "ISO3"
                             , nameJoinColumn = "ISO_code")

# draw map
mapParams <- mapCountryData(sPDF, nameColumnToPlot=input$frdm, 
                            mapTitle=paste(input$frdm, "Index" , input$yr, sep = " "),
                            catMethod=0:10,
                            colourPalette=clr,
                            addLegend=FALSE )

# draw legend
do.call( addMapLegend, 
         c(mapParams, 
           legendLabels="all", 
           legendWidth=0.5, 
           legendIntervals="page", 
           legendMar = 4 ))
}, width = 1000, height = 800)
```

## Distribution of Freedom by Region

The violin plots show that western countries enjoy more freedom and the scores 
are more uniform amongst the countries, as opposed to the Middle East with 
pronounced outliers. Since North America consist of only 2 countries (Canada and 
USA), there is not enough data to generate a violinplot.
```{r}
# Violin plots + boxplots per region
# Choose year, choose column
inputPanel(
  selectInput("yr1", label = "Select Year",
              choices = levels(hfi$year), selected = levels(hfi$year)[length(levels(hfi$year))]),
  
  selectInput("frdm1", label = "Select Area of freedom",
              choices = colnames(hfi_summary)[5:7] , selected = "Human Freedom")
)

renderPlot({
  hfiYR1 <- hfi[hfi$year==input$yr1,]
  hfi_summYR1 <- hfi_summary[hfi_summary$year==input$yr1,]
  
viobox <- ggplot(hfi_summYR1, aes(x=hfi_summYR1$region, y=hfi_summYR1[,input$frdm1], 
                              fill=hfi_summYR1$region)) +   
  geom_violin(trim=FALSE,show.legend = FALSE) + 
  geom_boxplot(width=0.1, fill="white") +
  labs(title=paste(input$frdm1,"by Region in", input$yr1, sep = " "),x="Region", y=input$frdm1) 

viobox + theme_light() + theme(axis.text.x=element_text(angle=45, hjust=1), 
                            plot.margin=margin(1, 1, 1, 2, "cm"),
                            plot.title = element_text(hjust = 0.5,size=22),axis.text=element_text(size=12)) 
}, width = 800, height = 500)
```

## Evolution of freedom

``` {r}
inputPanel(
  selectInput("country", label = "Select Country",
              choices = levels(hfi$countries), selected = "Belgium")

)

renderPlot({

summ_country <- hfi_summary[hfi_summary$countries==input$country,]

a<-round(mean(summ_country$`Personal Freedom`, digit=0))
b<-round(mean(summ_country$`Economical Freedom`, digit=0))
c<-round(mean(summ_country$`Human Freedom`, digit=0))

d<-summ_country$`Personal Freedom`[nrow(summ_country)]
e<-summ_country$`Economical Freedom`[nrow(summ_country)]
f<-summ_country$`Human Freedom`[nrow(summ_country)]

summ_country <- melt(summ_country)

p1 <- ggplot(summ_country, aes(summ_country$year,summ_country$value,
                          fill = summ_country$variable)) 

# graph evolution areas of freedom
p1 + geom_line(aes(color = summ_country$variable),group = summ_country$variable, size=2)+ 
  labs(title=paste("Evolution of Freedom in" , input$country, sep = " "),x="Year", y="Score") +
  guides(color=guide_legend("Legend")) + 
  scale_color_manual(values=clr[c(a,b,c)]) + 
  theme_light() + 
  theme(plot.margin=margin(1, 1, 1, 1, "cm"), 
        plot.title = element_text(hjust = 0.5,size=22),axis.text=element_text(size=12),legend.text=element_text(size=12),legend.title=element_text(size=14)) +
  annotate("text", x=.75, y=c(d,e,f),label = c("PF","EF","HF"), 
           color=clr[c(a,b,c)])
}, width = 800, height = 600)
```

## Details

``` {r}
inputPanel(
  selectInput("yr2", label = "Select Year",
              choices = levels(hfi$year), selected = levels(hfi$year)[length(levels(hfi$year))]),
  
  selectInput("country2", label = "Select Country",
              choices = levels(hfi$countries), selected = "Belgium")

)

renderPlot({
  
### for selected country
hfi_country <- hfi[hfi$countries==input$country2,]
hfi_countryYR<- hfi_country[hfi_country$year==input$yr2,]

details_country<-hfi_countryYR[,seperate_this]

for(i in 1:length(cols)){
  assign(cols[i], details_country[,grep(paste(cols[i], "_*", sep = ""), 
                                        colnames(details_country))])
}

  
country_data <- list()
for(i in 1:length(cols)) {
  country_data[[i]] <- eval(parse(text=cols[i]))
}
country_data<-melt(country_data)
country_data$variable<- factor(country_data$variable, levels = country_data$variable, labels = 
                           unlist(subtitles)) 
country_data$L1<- factor(country_data$L1, levels = 1:12, 
                   labels = titles)

p2 <-ggplot(country_data, value.name=titles,aes(x=variable,y=value, fill=value))

p2 + geom_col(position="dodge")+
  ylim(0,10) +
  coord_flip()+
  facet_wrap(~L1, nrow=12, scales = "free") + theme_light() +
  scale_fill_gradient2( midpoint=5, low=clr[5], mid=clr[7], high=clr[9],guide=FALSE) +
  labs(title=paste("Areas of Freedom for", input$country2, "in", input$yr2, 
                   sep = " "),x=element_blank(), y=element_blank()) +
  theme(plot.title = element_text(hjust = 0.5,size=22),axis.text=element_text(size=12),strip.text.x = element_text(size=14))

}, width = 800, height = 4000)  
``` 

## Explore different components of freedom

``` {r}
# Choose year, choose column
inputPanel(
  selectInput("yr3", label = "Select Year",
              choices = levels(hfi$year), selected = levels(hfi$year)[length(levels(hfi$year))]),
  
  selectInput("frdm3", label = "Select Area of freedom",
              choices = colnames(totals) , selected = 1)
)

renderPlot({
# Worldmap with totals for 2016

hfiYR3 <- hfi[hfi$year==input$yr3,]
totals<-hfiYR3[,!(colnames(hfi) %in% seperate_this)]
totals<-totals[,c(1,2,5:11,14:18)] # cut out pf and ef totals and rank
colnames(totals)<-c("year","ISO_code",titles)


# create SpatialPolygonsDataFrame: match countries between data and package
sPDF <- joinCountryData2Map( totals
                             , joinCode = "ISO3"
                             , nameJoinColumn = "ISO_code")

# draw map
mapParams <- mapCountryData(sPDF, nameColumnToPlot=input$frdm3, 
                            mapTitle=paste(input$frdm3 , input$yr3, sep = " "),
                            catMethod=0:10,
                            colourPalette=clr,
                            addLegend=FALSE )

# draw legend
do.call( addMapLegend, 
         c(mapParams, 
           legendLabels="all", 
           legendWidth=0.5, 
           legendIntervals="page", 
           legendMar = 4 ))
}, width = 1000, height = 800)
```